\textbf{Too Beautiful to be Fake: Attractive Faces are Less Likely to be Judged as Artificially Generated}

### 2022-09-21 submission/manuscript.Rmd
Update manuscript.Rmd
update results

  - id            : "1"
    institution   : "School of Social Sciences, Nanyang Technological University, Singapore"
  - id            : "2"
    institution   : "LKC Medicine, Nanyang Technological University, Singapore"
  - id            : "3"
    institution   : "National Institute of Education, Singapore"
  - id            : "4"
    institution   : "Centre for Research and Development in Learning, Nanyang Technological University, Singapore"
authornote: |
  Correspondence concerning this article should be addressed to Dominique Makowski, HSS 04-18, 48 Nanyang Avenue, Singapore (dom.makowski@gmail.com).
abstract: |
  Abstract abstract abstract.
keywords          : "visual illusions, illusion game, Pyllusion, personality, general factor"
wordcount         : "5114"
bibliography      : references.bib
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/apa.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
---

@@

In the first part of the study, participants answered a series of personality questionnaires, including the *Mini-IPIP6* [24 items, @sibley2011] measuring 6 personality traits, the *SIAS-6* and the *SPS-6* [6 items each, @peters2012development] assessing social anxiety levels, the *FFNI-BF* [30 items, @jauk2022validation] measuring 9 facets of narcissism; the *R-GPTS* [18 items, @freeman2021revised] measuring 2 dimensions related to paranoid thinking; and the *IUS-12* [12 items, @carleton2007fearing] measuring intolerance to uncertainty. Finally, we devised 5 items pertaining to expectations about AI-generated images technology ("I think current Artificial Intelligence algorithms can generate very realistic images"). To lower their saliency and the possibility of it priming the subjects about the task, we mixed these items with 5 items from the general attitudes towards AI scale [*GAAIS*, @schepman2020initial]. This scale was presented after the social anxiety questionnaires, and 3 attention check questions were embedded in the set of scales.

In the second part of this study, 109 photos of neutral-expression faces of real individuals from the validated American Multiracial Face Database (AMFD, [@chen2021broadening]) were presented to the participants for 500ms each, in a randomized order, following a fixation cross display for 750ms. After each stimulus presentation, ratings of **Attractiveness** ("I find this person attractive"), **Beauty** ("This face is good-looking"), **Trustworthiness** ("I find this person trustworthy") and **Familiarity** ("This person reminds me of someone I know") were collected using visual analog scales. Notably, as facial attractiveness is a multidimensional construct, encompassing evolutionary, sociocultural, biological as well as cognitive aspects [@rhodes2006evolutionary; @han2018beauty], we assessed attractiveness using 2 scales, measuring general attractiveness (**Attractiveness**) and physical attractiveness (**Beauty**), to determine whether the effects on simulation monitoring are driven by physical features.

In the last part of the study, participants were informed that about half of the face images previously seen were AI-generated (the instructions used a cover story mentioning that the research was aimed at validating a new face generation algorithm). The same set of stimuli was displayed again for 500 ms in a new randomized order. This time, after each display, participants were asked to express their belief regarding the nature of the stimulus using a visual analog scale (with *Fake* and *Real* as the two extremes). The study was implemented using *jsPsych* [@de2015jspsych], and the full set of instructions is available in the experiment code.

@@

One hundred and three participants were recruited via *Prolific*, the the crowd-sourcing platform providing the best data quality [@peer2022]. The only inclusion criterion was a fluent proficiency in English to ensure that the experiment instructions would be well-understood. Participants were incentivised with a reward of about \textsterling 7.5 for completing the study, which took about 45 minutes to finish. Demographic variables (age, gender, sexual orientation, education and ethnicity) were self-reported on a voluntary basis.

@@

The real-fake ratings (measured originally on a [-1, 1] analog scale) were converted into two scores, corresponding to two distinct mechanisms: the dichotomous *belief* (real or fake, derived based on the sign of the rating) and the *confidence* (the rating's absolute value) associated with that belief. Models predicting the former were set as logistic mixed models (with the participants and images entered as random factors), and models modeling the latter, as well as the other face ratings (attractiveness, beauty, trustworthiness and familiarity) were modeled using beta regression models (suited for an outcome variable expressed in percentages).

@@

The determinants of reality beliefs were modeled separately for attractiveness, beauty, trustworthiness, and familiarity, using second order raw polynomials coefficients to allow for possible quadratic relationships (**Figure 2**. Aside from attractiveness (conceptualized as a general construct), models for beauty, trustworthiness and familiarity were adjusted for the the two remaining variables *mutatis mutandis*. We took into account the gender of participants and stimuli by retaining the pictures that were aligned with the participants' sexual preference (e.g., female faces for homosexual females, male faces for heterosexual females, and both for bisexual participants), and modeling the interaction with the participants' gender. For the attractiveness and beauty models, we then added the interaction with the reported self-attractiveness (the average of the two questions pertaining to it) to investigate its potential modulatory effect.

```{r message=FALSE, warning=FALSE, fig.cap="Top part shows blabla.", out.width="100%"}
knitr::include_graphics("../figures/Figure2.png")
```

Finally, we investigated the inter-individual correlates of simulation monitoring by computing, for each participant, the proportion of faces judged as real (i.e., the overall bias towards one or the other belief), as well as the average confidence for faces judged as real, and fake. We assessed the link between these scores and dispositional traits using Bayesian correlation analysis [@makowski2019indices; @correlationArticle].

The analysis was carried out using *R 4.2* [@RCoreTeam2022], the *tidyverse* [@wickham2019], and the *easystats* collection of packages [@insightArticle; @bayestestRArticle; @performanceArticle; @parametersArticle]. As all the details, scripts and complimentary analyses are open-access, the manuscript will focus on significant results.

@@

Bayesian correlations with personality traits suggested that Honesty-Humility was negatively associated with the confidence in reality ($r = -0.21$, $95\% CI = [-0.38, -0.03]$, $BF_{10} = 3.57$), and positively associated with the Narcissism trait of Acclaim Seeking ($r = 0.26$, $95\% CI = [0.08, 0.43]$, $BF_{10} = 14.38$) and Grandiose Fantasies ($r = 0.22$, $95\% CI = [0.04, 0.40]$, $BF_{10} = 4.18$). Acclaim Seeking was also positively related with the confidence in fake judgments ($r = 0.22$, $95\% CI = [0.04, 0.40]$, $BF_{10} = 4.52$). No significant correlations was found for social anxiety, intolerance to uncertainty, or paranoid beliefs.

Questions pertaining to the attitude towards AI were reduced to 3 dimensions through factor analysis, labelled AI-Enthusiasm (loaded by items expressing interest and excitement in AI development and applications), AI-Realness (loaded by items expressing positive opinions on the ability of AI to create realistic material), and AI-Danger (loaded by items expressing concerns on the unethical misuse of AI technology). Only AI-Enthusiasm displayed a significant positive relationship with the confidence in both real ($r = 0.24$, $95\% CI = [0.06, 0.41]$, $BF_{10} = 8.00$) and fake ($r = 0.28$, $95\% CI = [0.11, 0.44]$, $BF_{10} = 23.04$) judgments.

@@

<!-- Attractiveness -->
Although attractiveness did not seem to be the primary drive underlying simulation monitoring of face images, we do nonetheless report significant associations, with a different pattern observed depending on the participant's gender. In particular, the quadratic relationship found for female participants is aligned with our hypothesis that salient faces (i.e., rated as very attractive or very unattractive) are judged to be more real. Alternatively, we could interpret the mostly positive linear relationship for males under an evolutionary lens. Specifically, males purportedly place more emphasis on facial attractiveness as a sign of reproductive potential, as compared with females, who tend to value characteristics signaling resource acquisition capabilities [@qi2022gender; @fink2006facial]. However, while it is possible that the evolutionary weight associated with attractiveness skewed the perceived saliency towards attractive faces, rendering attractive faces as significantly more salient than unattractive ones and in turn distorting the relationship with simulation monitoring, beauty, operationalized as physical attractiveness, was found to have a significant positive linear relationship with simulation monitoring decisions for males even after adjusting for the effects of trustworthiness and familiarity; both characteristics that are evolutionarily advantageous [@hou2019survival; @little2011facial]. As such, future studies are warranted to test this saliency-based hypothesis by measuring constructs closer to salience and its effects, for instance using neuroimaging [@lou2015perceptual; @indovina2007dissociation] or physiological markers [in particular, heart rate deceleration, @goldstein2002influence].

<!-- trustworthiness -->
In contrast to the findings for attractiveness, trustworthiness linearly increased beliefs in realness for females only. Given the evidence suggesting that faces presented as real were rated as more trustworthy [@liefooghe2022faces; @hoogers2021effect; @balas2017], we expected such link to be present for both genders. One of the contributing mechanism to this dimorphism could have been the increased risk-taking aversion reported in females [explained evolutionarily as a compromise to their reproductive potential, @van2020sex], to which perceived facial trustworthiness relates [@hou2019survival]. That being the case, however, faces judged as highly untrustworthy should have appeared as even more salient (representing an evolutionary threat), and hence be judged as more real. As such, further studies are needed to investigate the causes of the increased simulation monitoring sensitivity to trustworthiness in females.

<!-- familiarity -->
Contrary to our hypothesis, familiarity was not found to be significantly related to simulation monitoring decisions. However, it is to note that the distribution of familiarity ratings was strongly skewed, and that a low number of pictures was rated as highly familiar. Future studies could re-examine this aspect by experimentally manipulating familiarity, for instance by modulating the amount of exposures of items before the simulation monitoring ratings.

<!-- inter-indiv correlates -->
Unsurprisingly, participants with higher scores in Narcissism traits of Acclaim Seeking and Grandiose Fantasies were generally more confident in their simulation monitoring decisions. While early researchers had proposed narcissism to be a unitary construct [@raskin1988principal; @sedikides2004normal], recent studies cast narcissism as encompassing 2 unique dimensions - namely narcissistic vulnerability, which is characterized by low self-esteem and a need for admiration, and narcissistic grandiosity, which encompasses high self-esteem, arrogance and aggressive self-assertiveness [@pincus2011narcissistic]. Whereas narcissism has long been linked to over-confidence in decision-making [@campbell2004narcissism], a growing body of research indicates that this relationship is likely mostly driven by facets of narcissistic grandiosity (such as Acclaim Seeking), which drives individuals to see themselves as more superior than others and to externalize blame for poor decision-making consequences [@o2021grandiose]. Indeed, the evidence alludes that attributes specific to narcissistic grandiosity predict individuals' overconfidence in cognitive decision-making [@chatterjee2017master; @brunell2017narcissism; @o2021grandiose].

@@

We would like to thank **TODO: add STUDENT NAME** for his contribution to the selection of the materials.

>>>>

  - id            : "1"
    institution   : "School of Social Sciences, Nanyang Technological University, Singapore"
  - id            : "2"
    institution   : "LKC Medicine, Nanyang Technological University, Singapore"
  - id            : "3"
    institution   : "National Institute of Education, Singapore"
  - id            : "4"
    institution   : "Centre for Research and Development in Learning, Nanyang Technological University, Singapore"
authornote: |
  Correspondence concerning this article should be addressed to Dominique Makowski, HSS 04-18, 48 Nanyang Avenue, Singapore (dom.makowski@gmail.com).
abstract: |
  Abstract abstract abstract.
keywords          : "attractiveness, simulation monitoring, fiction, deep fakes, sense of reality"
wordcount         : "5114"
bibliography      : references.bib
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/apa.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
---

@@

In the first part of the study, participants answered a series of personality questionnaires, including the *Mini-IPIP6* [24 items, @sibley2011] measuring 6 personality traits, the *SIAS-6* and the *SPS-6* [6 items each, @peters2012development] assessing social anxiety levels, the *FFNI-BF* [30 items, @jauk2022validation] measuring 9 facets of narcissism; the *R-GPTS* [18 items, @freeman2021revised] measuring 2 dimensions related to paranoid thinking; and the *IUS-12* [12 items, @carleton2007fearing] measuring intolerance to uncertainty. Finally, we devised 5 items pertaining to expectations about AI-generated images technology ("I think current Artificial Intelligence algorithms can generate very realistic images"). To lower their saliency and the possibility of it priming the subjects about the task, we mixed these items with 5 items from the general attitudes towards AI scale [*GAAIS*, @schepman2020initial]. This scale was presented after the social anxiety questionnaires, and 3 attention check questions were embedded in the surveys.

In the second part of this study, 109 photos of neutral-expression faces of real individuals from the validated American Multiracial Face Database (AMFD, [@chen2021broadening]) were presented to the participants for 50 0ms each, in a randomized order, following a fixation cross display (750 ms). After each stimulus presentation, ratings of *Trustworthiness* ("I find this person trustworthy") and *Familiarity* ("This person reminds me of someone I know") were collected using visual analog scales. Notably, as facial attractiveness is a multidimensional construct, encompassing evolutionary, sociocultural, biological as well as cognitive aspects [@rhodes2006evolutionary; @han2018beauty], we assessed attractiveness using 2 scales, measuring general *Attractiveness* ("I find this person attractive") and physical *Beauty* ("This face is good-looking").

In the last part of the study, participants were informed that about half of the face images previously seen were AI-generated (the instructions used a cover story mentioning that the research was aimed at validating a new face generation algorithm). The same set of stimuli was displayed again for 500 ms in a new randomized order. This time, after each display, participants were asked to express their belief regarding the nature of the stimulus using a visual analog scale (with *Fake* and *Real* as the two extremes). The study was implemented using *jsPsych* [@de2015jspsych], and the exact instructions are available in the experiment code.

@@

One hundred and three participants were recruited via *Prolific*, a crowd-sourcing platform recognized for providing high quality data [@peer2022]. The only inclusion criterion was a fluent proficiency in English to ensure that the experiment instructions would be well-understood. Participants were incentivised with a reward of about \textsterling 7.5 for completing the study, which took about 45 minutes to finish. Demographic variables (age, gender, sexual orientation, education and ethnicity) were self-reported on a voluntary basis.

@@

The real-fake ratings (measured originally on a [-1, 1] analog scale) were converted into two scores, corresponding to two conceptually distinct mechanisms: the dichotomous *belief* (real or fake, derived based on the sign of the rating) and the *confidence* (the rating's absolute value) associated with that belief. The former was analyzed using logistic mixed models (with the participants and images entered as random factors), which models probability of assigning a face to the real (>= 0) as opposed to fake (< 0). The latter, as well as the other face ratings (attractiveness, beauty, trustworthiness and familiarity) was modeled using beta regression models (suited for outcome variables expressed in percentages).

@@

The determinants of reality beliefs were modeled separately for attractiveness, beauty, trustworthiness, and familiarity, using second order raw polynomials coefficients to allow for possible quadratic relationships (**Figure 2**. Aside from attractiveness (conceptualized as a general construct), models for beauty, trustworthiness and familiarity were adjusted for the the two remaining variables *mutatis mutandis*. We took into account the gender of participants and stimuli by retaining the pictures that were aligned with the participants' sexual preference (e.g., female faces for homosexual females, male faces for heterosexual females, and both for bisexual participants), and modeling the interaction with the participants' gender. For the attractiveness and beauty models, we then added the interaction with the reported self-attractiveness (the average of the two questions pertaining to it) to investigate its potential modulatory effect. Finally, we investigated the inter-individual correlates of simulation monitoring with similar models (but this time, for all items regardless of the participant's gender or sexual orientation) for each questionnaire, with all of the subscales as orthogonal predictors.

```{r message=FALSE, warning=FALSE, fig.cap="Top part shows the efffect of face ratings on 1) the probability of judging a face as real vs. fake (solid line) and 2) on the confidence associated with that judgment (dashed lines) depending on the sex. Bottom part shows the effect of personality traits on the belief (black line) and the confidence associated with it (colored lines). The points are the average per participant confidence for both types of judgments. Stars indicate significance (p < .001***, p < .01**, p < .05*)", out.width="100%"}
knitr::include_graphics("../figures/Figure2.png")
```

The analysis was carried out using *R 4.2* [@RCoreTeam2022], the *tidyverse* [@wickham2019], and the *easystats* collection of packages [@insightArticle; @bayestestRArticle; @performanceArticle; @parametersArticle; @correlationArticle]. As all the details, scripts and complimentary analyses are open-access, the manuscript will focus on significant results.

@@

The models including the personality traits suggested that *Honesty-Humility* had a significant negative relationship predominantly with the confidence associated with faces judged as real ($\beta_{real} = -1.69$, $95\% CI = [-2.74, -0.63]$, $z = -3.14$, $p < .01$; $\beta_{fake} = -1.02$, $95\% CI = [-2.07, 0.03]$, $z = -1.90$, $p = 0.058$). Narcissism shared significant links through various dimensions, such as positive relationship between the confidence for both real and fake judgments and *Acclaim Seeking* ($\beta_{real} = 2.08$, $95\% CI = [0.79, 3.37]$, $z = 3.16$, $p < .01$; $\beta_{fake} = 1.60$, $95\% CI = [0.30, 2.90]$, $z = 2.42$, $p < .05$), and a negative relationship with *Authoritativeness* ($\beta_{real} = -1.38$, $95\% CI = [-2.50, -0.27]$, $z = -2.43$, $p < .05$; $\beta_{fake} = -1.23$, $95\% CI = [-2.35, -0.11]$, $z = -2.15$, $p < .05$) and *Thrill Seeking* ($\beta_{real} = -0.93$, $95\% CI = [-1.75, -0.10]$, $z = -2.20$, $p < .05$; $\beta_{fake} = -1.03$, $95\% CI = [-1.87, -0.20]$, $z = -2.44$, $p < .05$). Finally, the social phobia dimension was significantly associated with a decreased probability of judging faces as fake ($\beta = -0.97$, $95\% CI = [-1.86, -0.08]$, $z = -2.14$, $p < .05$), but no effect on the confidence judgments was found. No significant relationships was found for intolerance to uncertainty or paranoid beliefs.

<!-- Bayesian correlations with personality traits suggested that Honesty-Humility was negatively associated with the confidence in reality ($r = -0.21$, $95\% CI = [-0.38, -0.03]$, $BF_{10} = 3.57$), and positively associated with the Narcissism trait of Acclaim Seeking ($r = 0.26$, $95\% CI = [0.08, 0.43]$, $BF_{10} = 14.38$) and Grandiose Fantasies ($r = 0.22$, $95\% CI = [0.04, 0.40]$, $BF_{10} = 4.18$). Acclaim Seeking was also positively related with the confidence in fake judgments ($r = 0.22$, $95\% CI = [0.04, 0.40]$, $BF_{10} = 4.52$). No significant correlations was found for social anxiety, intolerance to uncertainty, or paranoid beliefs. -->

Questions pertaining to the attitude towards AI were reduced to 3 dimensions through factor analysis, labelled AI-Enthusiasm (loaded by items expressing interest and excitement in AI development and applications), AI-Realness (loaded by items expressing positive opinions on the ability of AI to create realistic material), and AI-Danger (loaded by items expressing concerns on the unethical misuse of AI technology). Only AI-Enthusiasm displayed a significant positive relationship with the confidence in both real and fake judgments ($\beta_{real} = 0.29$, $95\% CI = [0.08, 0.50]$, $z = 2.74$, $p < .01$; $\beta_{fake} = 0.38$, $95\% CI = [0.18, 0.59]$, $z = 3.60$, $p < 0.001$).

<!-- Only AI-Enthusiasm displayed a significant positive relationship with the confidence in both real ($r = 0.24$, $95\% CI = [0.06, 0.41]$, $BF_{10} = 8.00$) and fake ($r = 0.28$, $95\% CI = [0.11, 0.44]$, $BF_{10} = 23.04$) judgments. -->

@@

<!-- Attractiveness: females -->
Although attractiveness did not seem to be the primary drive underlying simulation monitoring of face images, we do nonetheless report significant associations, with a different pattern observed depending on the participant's gender. The quadratic relationship found for female participants is aligned with our hypothesis that salient faces (i.e., rated as very attractive or very unattractive) are judged to be more real. The fact that this effect did not reach significance for beauty underlines that attractiveness judgments, and its role in simulation monitoring, are a multidimensional construct that cannot be reduced to physical facial attractiveness, in particular for women **[REF?]**.

<!-- Attractiveness: males -->
Indeed, we found a significant positive linear relationship in male participants for both attractiveness and beauty on simulation monitoring, that we could interpret under an evolutionary lens. Specifically, males purportedly place more emphasis on facial attractiveness as a sign of reproductive potential, as compared with females, who tend to value characteristics signaling resource acquisition capabilities [@qi2022gender; @fink2006facial]. However, while it is possible that the evolutionary weight associated with attractiveness skewed the perceived saliency towards attractive faces, rendering them significantly more salient than unattractive faces, in turn distorting the relationship with simulation monitoring. However, future studies should test this saliency-based hypothesis by measuring constructs closer to salience and its effects, for instance using neuroimaging [@lou2015perceptual; @indovina2007dissociation] or physiological markers [in particular, heart rate deceleration, @goldstein2002influence].

<!-- even after adjusting for the effects of trustworthiness and familiarity; both characteristics that are evolutionarily advantageous [@hou2019survival; @little2011facial]. A -->

<!-- trustworthiness -->
Our results suggested that trustworthiness ratings linearly increased beliefs in realness, for females only. Given the evidence suggesting that faces presented as real were rated as more trustworthy [@liefooghe2022faces; @hoogers2021effect; @balas2017], we expected such link to be present for both genders. One of the contributing mechanism to this dimorphism could have been the increased risk-taking aversion reported in females [explained evolutionarily as a compromise to their reproductive potential, @van2020sex], to which perceived facial trustworthiness relates [@hou2019survival]. However, if that was the case, faces judged as highly untrustworthy should have appeared as even more salient (representing an evolutionary threat), and hence be judged as more real. As such, further studies are needed to investigate the causes of the increased simulation monitoring sensitivity to trustworthiness in females.

<!-- familiarity -->
Contrary to our hypothesis, familiarity was not found to be significantly related to simulation monitoring decisions. However, the distribution of familiarity ratings was strongly skewed, and only a low number of pictures was rated as highly familiar. Future studies could re-examine this aspect by experimentally manipulating familiarity, for instance by modulating the amount of exposure to items before the simulation monitoring judgments.

<!-- inter-indiv correlates -->
Regarding the role of inter-individual characteristics in simulation monitoring tendencies, we found that participants with stronger narcissism (in particular the acclaim seeking and grandiose fantasies dimensions) displayed a stronger confidence in their simulation monitoring judgments. This is in line with the literature

While early researchers had proposed narcissism to be a unitary construct [@raskin1988principal; @sedikides2004normal], recent studies cast narcissism as encompassing 2 unique dimensions - namely narcissistic vulnerability, which is characterized by low self-esteem and a need for admiration, and narcissistic grandiosity, which encompasses high self-esteem, arrogance and aggressive self-assertiveness [@pincus2011narcissistic]. Whereas narcissism has long been linked to over-confidence in decision-making [@campbell2004narcissism], a growing body of research indicates that this relationship is likely mostly driven by facets of narcissistic grandiosity (such as Acclaim Seeking), which drives individuals to see themselves as more superior than others and to externalize blame for poor decision-making consequences [@o2021grandiose]. Indeed, the evidence alludes that attributes specific to narcissistic grandiosity predict individuals' overconfidence in cognitive decision-making [@chatterjee2017master; @brunell2017narcissism; @o2021grandiose].

@@

We would like to thank Taong Ren Qing Malcolm for his contribution to the selection of the materials.
